Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2024-04-18 src/tex/ms.tex
remove highlight

\begin{abstract}
	We investigate the possibility of improving the accuracy of the
	phenomenological waveform model, IMRPhenomD, by jointly optimizing all the
	calibration coefficients at once, given a set of numerical relativity (NR)
	waveforms. When IMRPhenomD was first calibrated to NR waveforms, different
	parts (i.e., the inspiral, merger, and ringdown) of the waveform were 
	calibrated separately. 	Using \ripple, a library of waveform models 
	compatible with automatic differentiation, we can, for the first time, 
	perform gradient-based optimization on all the waveform coefficients at the 
	same time. This joint optimization process allows us to capture previously 
	ignored correlations between separate parts of the waveform. We found that 
	after recalibration, the median mismatch between the model and NR waveforms 
	decreases by $50\%$. We further explore how different regions of the source 
	parameter space respond to the optimization procedure. We find that the 
	degree of improvement correlates with the spins of the source. This work 
	shows a promising avenue to help understand and treat systematic error 
	in waveform models.
\end{abstract}

@@

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective/loss function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of the optimized waveforms with the original waveforms. We also show how the
optimization result differs as a function of the source parameters. Finally, in
Sec.~\ref{sec:discussion}, we discuss the difference between our calibration
procedure and the procedure used in \citep{khan2016frequency}. We also explain how the reduced spin
parameterization affects the accuracy of the model. 
Note that throughout this paper we use the terms recalibration and optimization interchangeably.

@@

We note that the parameter space we use is restricted to $q\leq8$, while waveforms 
of $q=18$ are used in the original calibration trial. Also, we have directly used NR 
waveforms instead of hybridized SEOB-NR waveforms for training and validation purposes. 
These choices put the original IMRPhenomD result at a disadvantage, as both the parameter 
space and frequency range are restricted. We tested our result by training the model 
using a set of training waveforms of the same intrinsic parameters, but with BAM waveforms 
replaced with SEOBNR waveforms. We can see the result in Fig.~\ref{fig:q14818}, where the 
mismatch median decreases by about $8\%$. Nevertheless, it is known that IMRPhenomD has a large 
inaccuracy in the large mass-ratio regime, and is rarely used in this region of parameter 
space. Hence, we focus on showing that AD with gradient descent, as a general and robust method, 
can be used to further optimize GW waveform models with arbitrary sets of training 
waveforms used.

>>>>

\begin{abstract}
	We investigate the possibility of improving the accuracy of the
	phenomenological waveform model, IMRPhenomD, by jointly optimizing all the
	calibration coefficients at once, given a set of numerical relativity (NR)
	waveforms. When IMRPhenomD was first calibrated to NR waveforms, different
	parts (i.e., the inspiral, merger, and ringdown) of the waveform were 
	calibrated separately. Using \ripple, a library of waveform models 
	compatible with automatic differentiation, we can, for the first time, 
	perform gradient-based optimization on all the waveform coefficients at the 
	same time. This joint optimization process allows us to capture previously 
	ignored correlations between separate parts of the waveform. We found 
	that after recalibrating with a slightly restricted parameter subspace 
	$(q\leq8)$, despite the tail of the mismatch distribution remains similar, 
	the median mismatch between the model and NR waveforms decreases by 
	$50\%$. We further explore how different regions of the source 
	parameter space respond to the optimization procedure. We find that the 
	degree of improvement correlates with the spins of the source. This work 
	shows a promising avenue to help understand and treat systematic error 
	in waveform models.
\end{abstract}

@@

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective/loss function for the calibration, followed by 
outlining the specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of the optimized waveforms with the original waveforms. We also show how the
optimization result differs as a function of the source parameters. Finally, in
Sec.~\ref{sec:discussion}, we discuss the differences between our calibration
procedure and the procedure used in \citep{khan2016frequency}. Additionally, 
we discuss some caveats (such as the difference in training waveforms) that limit 
our conclusions on the mismatch improvement. We also explain how the reduced spin 
parameterization affects the accuracy of the model. Note that throughout this 
paper we use the terms recalibration and optimization interchangeably.

@@

We note that the parameter space we use is restricted to $q\leq8$, while waveforms 
of $q=18$ are used in the original calibration trial. Also, we have directly used NR 
waveforms instead of hybridized SEOB-NR waveforms for training and validation purposes. 
These choices may put the original IMRPhenomD result at a disadvantage when 
assessing accuracy, as both the parameter space and frequency range are restricted. We 
tested our result by training the model using a set of training waveforms of the same 
intrinsic parameters, but with BAM waveforms replaced with SEOBNR waveforms. We can see 
the result in Fig.~\ref{fig:q14818}, where the mismatch median decreases by about $8\%$. 
Nevertheless, it is known that IMRPhenomD has a large inaccuracy in the large mass-ratio 
regime, and is rarely used in this region of parameter space. Hence, we focus on showing 
that AD with gradient descent, as a general and robust method, can be used to further 
optimize GW waveform models with arbitrary sets of training waveforms used. With 
more publicly available data in the future, more systematic studies should be made to 
precisely assess the improvement one can get out of further optimization.
